Supplementary Materials for CBEA: Competitive Balances for Taxonomic Enrichment Analysis

### 2022-02-15 supplemental.tex
Final manuscript

Fig~\ref{fig:s5} shows results for simulation studies as detailed in the \nameref{methods} section. Panel A presents results for a regression learning task with a continuous outcome. We observed no difference in performance values across different evaluation models. Overall, prediction error did not change across sparsity levels, and decreases with increasing signal-to-noise ratio (SNR). However, at higher correlation levels, the pattern was more erratic. For example, when $\rho$ is set at $0.5$, higher SNR decreases performance only at low sparsity $p = 0.2$, but had the expected pattern medium sparsity $p = 0.4$, where higher SNR correlates with improved performance. This effect seems to be specific to low effect saturation scenarios (only a limited number of taxa sets are associated with the outcome). Interestingly, higher sparsity levels produces better performance.

@@

\clearpage
\bibliography{references}{}
\bibliographystyle{plos2015}

>>>>

Fig~\ref{fig:s5} shows results for simulation studies as detailed in the Methods section. Panel A presents results for a regression learning task with a continuous outcome. We observed no difference in performance values across different evaluation models. Overall, prediction error did not change across sparsity levels, and decreases with increasing signal-to-noise ratio (SNR). However, at higher correlation levels, the pattern was more erratic. For example, when $\rho$ is set at $0.5$, higher SNR decreases performance only at low sparsity $p = 0.2$, but had the expected pattern medium sparsity $p = 0.4$, where higher SNR correlates with improved performance. This effect seems to be specific to low effect saturation scenarios (only a limited number of taxa sets are associated with the outcome). Interestingly, higher sparsity levels produces better performance.

@@

\section{Single sample type I error diagnostic plot}
\begin{figure}[!h]
    \centering
    \includegraphics[width=0.9\textwidth]{figures/data_ss_diag_ridges.png}
    \caption{\textbf{Distribution of type I error values across app relications in real data random set evaluations for CBEA inference at the sample level}. Density (\emph{y}-axis) for type I error values (\emph{x}-axis) of each evaluated approach for sample-level inference using real data across 500 replications. Here, type I error was estimated as the proportion of samples where a randomly sampled set of different sizes where identified to be statistically significant at $p$-value threshold of 0.05} 
    \label{fig:s7}
\end{figure}

\clearpage
%\bibliography{references}{}
\begin{thebibliography}{10}

\bibitem{delignette-muller2015}
{Delignette-Muller} ML, Dutang C.
\newblock Fitdistrplus: {{An R}} Package for Fitting Distributions.
\newblock Journal of Statistical Software. 2015;64(4):1--34.

\bibitem{benaglia2009}
Benaglia T, Chauveau D, Hunter DR, Young D.
\newblock Mixtools: {{An R}} Package for Analyzing Finite Mixture Models.
\newblock Journal of Statistical Software. 2009;32(6):1--29.

\bibitem{wu2012}
Wu D, Smyth GK.
\newblock Camera: A Competitive Gene Set Test Accounting for Inter-Gene
  Correlation.
\newblock Nucleic Acids Research. 2012;40(17):e133.
\newblock doi:{10.1093/nar/gks461}.

\bibitem{calgaro2020}
Calgaro M, Romualdi C, Waldron L, Risso D, Vitulo N.
\newblock Assessment of Statistical Methods from Single Cell, Bulk {{RNA-seq}},
  and Metagenomics Applied to Microbiome Data.
\newblock Genome Biology. 2020;21(1):191.
\newblock doi:{10.1186/s13059-020-02104-1}.

\bibitem{cario1997}
Cario MC.
\newblock Modeling and {{Generating Random Vectors}} with {{Arbitrary Marginal
  Distributions}} and {{Correlation Matrix}}; 1997.

\bibitem{agresti1998}
Agresti A, Coull BA.
\newblock Approximate {{Is Better}} than "{{Exact}}" for {{Interval
  Estimation}} of {{Binomial Proportions}}.
\newblock The American Statistician. 1998;52(2):119--126.
\newblock doi:{10.2307/2685469}.

\bibitem{frost2020}
Frost HR.
\newblock Variance-Adjusted {{Mahalanobis}} ({{VAM}}): A Fast and Accurate
  Method for Cell-Specific Gene Set Scoring.
\newblock Nucleic Acids Research. 2020;48(16):e94--e94.
\newblock doi:{10.1093/nar/gkaa582}.

\bibitem{delong1988}
DeLong ER, DeLong DM, {Clarke-Pearson} DL.
\newblock Comparing the Areas under Two or More Correlated Receiver Operating
  Characteristic Curves: A Nonparametric Approach.
\newblock Biometrics. 1988;44(3):837--845.

\bibitem{hawinkel2019}
Hawinkel S, Mattiello F, Bijnens L, Thas O.
\newblock A Broken Promise: Microbiome Differential Abundance Methods Do Not
  Control the False Discovery Rate.
\newblock Briefings in Bioinformatics. 2019;20(1):210--221.
\newblock doi:{10.1093/bib/bbx104}.

\bibitem{breiman2001}
Breiman L.
\newblock Random {{Forests}}.
\newblock Machine Learning. 2001;45(1):5--32.
\newblock doi:{10.1023/A:1010933404324}.

\bibitem{kuhn2020}
Kuhn M, Wickham H. Tidymodels: A Collection of Packages for Modeling and
  Machine Learning Using Tidyverse Principles.; 2020.

\bibitem{xiao2018}
Xiao J, Chen L, Yu Y, Zhang X, Chen J.
\newblock A {{Phylogeny-Regularized Sparse Regression Model}} for {{Predictive
  Modeling}} of {{Microbial Community Data}}.
\newblock Frontiers in Microbiology. 2018;9.
\newblock doi:{10.3389/fmicb.2018.03112}.

\bibitem{dong2020}
Dong M, Li L, Chen M, Kusalik A, Xu W.
\newblock Predictive Analysis Methods for Human Microbiome Data with
  Application to {{Parkinson}}'s Disease.
\newblock PLOS ONE. 2020;15(8):e0237779.
\newblock doi:{10.1371/journal.pone.0237779}.

\bibitem{hanzelmann2013}
H{\"a}nzelmann S, Castelo R, Guinney J.
\newblock {{GSVA}}: Gene Set Variation Analysis for Microarray and {{RNA-Seq}}
  Data.
\newblock BMC Bioinformatics. 2013;14(1):7.
\newblock doi:{10.1186/1471-2105-14-7}.

\bibitem{love2014}
Love MI, Huber W, Anders S.
\newblock Moderated Estimation of Fold Change and Dispersion for {{RNA-seq}}
  Data with {{DESeq2}}.
\newblock Genome Biology. 2014;15(12):550.
\newblock doi:{10.1186/s13059-014-0550-8}.

\bibitem{martin2020}
Martin BD, Witten D, Willis AD.
\newblock Modeling Microbial Abundances and Dysbiosis with Beta-Binomial
  Regression.
\newblock The Annals of Applied Statistics. 2020;14(1):94--115.
\newblock doi:{10.1214/19-AOAS1283}.

\end{thebibliography}

\bibliographystyle{plos2015}
